Towards automated provenance collection for experimental runs of agent-based models\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}

### 2023-04-19 paper/paper.tex
Updates from Overleaf

\begin{document}
%
\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{Supported by The Scottish Government}}
%
\titlerunning{An automated provenance collection framework}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Doug Salt\inst{1}\orcidID{0000-1111-2222-3333} \and
Gary Polhill\inst{2,3}\orcidID{1111-2222-3333-4444} }
%
\authorrunning{Salt et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{The James Hutton Institute, Craigiebuckler Aberdeen AB15 8QH Scotland
\email{doug.salt@hutton.ac.uk}\\
\url{http://hutton.ac.uk}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.

@@

Replication of social simulation results has been highlighted as a significant
issue for the ABM for a number of years (e.g.  \cite{edmonds2003replication}),
and in particular the paper that forms the basis of this work
\cite{polhill2017lessons}.  The focus of replication work has previously just
been on the model itself, but as was shown in ibid, the analysis of the
outputs of the model can potentially be just as complex, and no less difficult
to replicate unless adequate records are kept. Indeed there as an additional
attempt to reproduce the results and even with the lessons learnt from ibid, it
still took months to recreate the final diagrams that were submitted to paper.
The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook. There are also appear to be
many scientific workflow tools, examples being, Snakemake
\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
\cite{ludascher2006scientific}, Taverna \cite{hull2006taverna} just to name a
few. However these are workflow tools, the concentrate on automation and
repeatability, and if they do record provenance, then it is an after thought.
It appears that these tools are more centred on automation rather than
provenance and metadata.

One of the lessons learned from the replication exercise in
\cite{polhill2017lessons} show that, for the purposes of replicability more
detailed guidance on the information that should be recorded,  and on tools
that could be used to support the process. Ideally the aim should be to
completely automate the process, record accurately sources of data used,
version check applications used essentially providing a complete graph from
data to result.

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity
\cite{polhill_nonlinearities_2013,gimona2011exploring}. Belonging to the
‘typification’ class of social simulations  \cite{boero2005does}.  This work
involved the analysis of the outputs of approximately 20,000 runs of the model
using a number of techniques aimed at demonstrating non-linearities in the
relationship between incentivisation and biodiversity outcome.  Recording
workflow data on the process used to create analysis can be challenging, and
currently there are no codified standards as to how this should be done for
agent-based models. For FEARLUS-SPOMM, the methods used drew heavily on statistical
techniques available as R packages that are as part of core R functionality.
Although R allows transcripts of interactive terminal sessions to be saved, the
work involved great deal of exploration of different ways of attempting to
visualise and analyse the non-linearities in the model results, not all of
which were likely to be reported in the paper. Such logs are therefore not the
best way to record the means by which the outputs were analysed, and hence the
strategy used was to save each analysis or visualisation in a(n R) script.
Since the output from the (Swarm) software that generated the output data being
analysed used a mixture of text formats, some Perl scripts were also written to
process that output into a CSV file for easy import into R. When the MIRACLE
project \cite{parker2019final} provided a context in which the replication of
that analysis was necessary, an opportunity was created to test the viability
of the above strategy.

If we make the assumptions that the input data is accurate and relevant; the
model bug-free and rigorously correct then the manifest problem with
agent-based modelling is the number of replications required to provide
adequate probability that an interesting result is not the null hypothesis, and
just down to a matter of chance. Realistically this involves many, many
repetitions of the same experiment or small variations on the parameters for a
given experiment. Similarly infeasible is the manual running of these many
experiments and thus most resort to some kind of automation. We refer to this
as the automatic and replicability problem. In  addition to this there is the
problem of recording which data is being used, which data is being produced,
how it was gathered or created and how the resultant data was processed. Given
20,000 experiments the amount of such data is not inconsiderable. Moreover this
data needs to be in a form that is analysable and visualisable.

>>>>

\begin{document}
%
\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
%%% GP: if you ever want the standard acknowledgement text for C5, it's on our
%%% GP: project webpage at https://large-scale-modelling.hutton.ac.uk/
%
\titlerunning{An automated provenance collection framework}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }

%%% GP: to discuss -- which other authors should we include here?
%%% GP: Kit Macleod? (Has he had intellectual input into this via WP1?)
%%% GP: Coran Musk? (Random acknowledgement for looking after the machines)
%%% GP: Lorenzo Milazzo? (As the original SSREPI developer)
%%% GP: Dawn Parker? (As MIRACLE PI -- do we see this as a MIRACLE paper?)
%
\authorrunning{D. Salt et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
\email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
\url{https://www.hutton.ac.uk/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.

@@

Replication of social simulation results has been highlighted as a significant
issue for the ABM community for a number of years (e.g. \cite{edmonds2003replication}).
The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
the analysis of the outputs of the model can potentially be just as complex a
process as developing and running the model itself. Analysis of outputs is no less difficult
to replicate unless adequate records are kept. The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook. There are also 
many scientific workflow tools, such as Snakemake
\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
\cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
these are workflow tools, the focus is on automation and
repeatability rather than provenance, which, if it is included, is as an afterthought.

One of the lessons learned from the replication exercise in
\cite{polhill2017lessons} was that, for the purposes of replication, more
detailed guidance on the information that should be recorded is needed. Since
recording such metadata is tedious (and error-prone) for humans, any such
guidance should be accompanied by specification of tools
that could be used to support the process. In the ideal world, the process
of recording provenance metadata would be completely automated, essentially providing a complete graph from data (including their sources) through applications (model, scripts and analysis tools, including versions thereof) to result.

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity
\cite{polhill_nonlinearities_2013,gimona2011exploring}. Belonging to the
‘typification’ class of social simulations  \cite{boero2005does}, this work
involved the analysis of the outputs of approximately 20,000 runs of the model
using a number of techniques aimed at demonstrating nonlinearities in the
relationship between incentivisation and biodiversity outcome.  Recording
provenance metadata on the process used to analyse and visualise the outputs is challenging, and
currently there are no codified standards as to how this should be done for
agent-based models. For FEARLUS-SPOMM, the analysis and visualisation methods used drew heavily on statistical
techniques available as R packages.
Although R allows transcripts of interactive terminal sessions to be saved, the
work involved great deal of exploration of alternative analyses and visualisations, not all of
which were reported in the manuscript as finally accepted. Such transcripts are therefore not the
best way to record the means by which a model's outputs were analysed, and hence the
strategy used was to save each analysis or visualisation in a(n R) script.
Since the output from the (Swarm) model software used a mixture of text formats, some Perl scripts were also written to
process that output into CSV format for easy import into R. When the MIRACLE
project \cite{parker2019final} provided a context in which the replication of
that analysis was necessary, an opportunity was created to test the viability
of the strategy of relying on scripts to record provenance.

`Multifinality' (the same initial conditions and parameter settings having qualitatively different results) in agent-based models means we need to ensure that
reported results are not just down to a matter of chance. This is one of many reasons (e.g. in empirical contexts especially, calibration, validation and sensitivity analysis) why experiments with ABMs involve large numbers of runs. Most ABM experiments therefore resort to some kind of automation, including of the kind provided by workflow tools mentioned above, but also using built-in features of ABM software (e.g. in the case of NetLogo, BehaviorSpace). We refer to this
as the automatic and replicability problem. In addition to this there is the
problem of recording which data is being used, which data is being produced,
how it was gathered or created and how the resultant data was processed. Given
20,000 experiments the amount of such data is not inconsiderable. Moreover this
data needs to be in a form that is analysable and visualisable.
